---
title: "<center><h1>Image analysis with EBImage</h1></center>"
#author: "Rossella Tufano"
#date: "28/12/2018"
output:
   html_document:
    toc: yes
---

<style>
//  h1, h2, h3, h4, a, table {
//    font-family: "Times New Roman";
//  }

    h1 {
      font-weight:bold;
    }
    
    h1, h2, h3, h4, a {
        color: #448FA7;
    }
    
    table, td {
      border: 1px solid black;
      text-align: "center";
      vertical-align: "middle";
      padding: 7px;
    }
    
    #kable.wrapper{
      border: none;
    }
</style>

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

[EBImage](https://www.bioconductor.org/packages/release/bioc/html/EBImage.html) is a Bioconductor package that offers tools to process and analyze images. 

## Installation

To install EBImage, start R (version "3.5") and enter:

```{r eval=FALSE}
#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("EBImage", version = "3.8")
```

## Loading 

To load EBImage, use the following command:

```{r}
library("EBImage")
```

## Reading

Images are loaded using the function `readImage()`, which takes as input a character vector of file names or URLs. The loaded images are stored into an Image object.

```{r}
#files=list.files(system.file("images",package="EBImage"))
#sampleColor=readImage(system.file("images","sample-color.png",package="EBImage"))
#sample=readImage(system.file("images","sample.png",package="EBImage"))
#shapes=readImage(system.file("images","shapes.png",package="EBImage"))
vesuvio=readImage("vesuvio.png")
cells=readImage(system.file("images","cells.tif",package="EBImage"))
nuclei=readImage(system.file("images","nuclei.tif",package="EBImage"))
greyVesuvio=channel(vesuvio,'grey')
```

### Grayscale images

```{r}
str(greyVesuvio)
```

```{r}
print(greyVesuvio)
```

```{r}
dim(greyVesuvio)
```

```{r}
imageData(greyVesuvio)[10:14,4:8]
```

### Color images

```{r}
print(vesuvio,short=TRUE)
```

### Image stacks

```{r}
print(nuclei,short=TRUE)
```

An image histogram shows the distribution of pixel intensities.

```{r}
hist(greyVesuvio)
range(greyVesuvio)
```

## Displaying

The function `display()` allows to display images using an interactive JavaScript viewer or R’s built-in graphics capabilities.

```{r}
display(greyVesuvio)
display(vesuvio)
display(nuclei,all=TRUE)
```

## Writing

```{r}
writeImage(greyVesuvio,'vesuvio.jpeg',quality = 85)
file.size('vesuvio.jpeg')
```

## Image processing

The operations that can be applied to an image can be classified into three categories: **point, local and global**.

**Point operations**: output image pixel value at a specific position depends only on the input image pixel value at the same position. Examples: brigthness and contrast adjustment, gamma correction, thresholding.

**Local operations**: output image pixel value at a specific position depends on the pixel values in the neighborhood of input image pixel value at the same position. Example: convolution.

**Global operations**: output image pixel value at a specific position depends on all pixel values in the input image. 

### Brightness adjustment

##### $y=x+b$

```{r}
display(
   combine(
     vesuvio+0.2,
     vesuvio,
     vesuvio-0.2
   ), all=TRUE
 )
``` 

### Contrast adjustment

##### $y=x*c$

```{r}
display(
   combine(
     vesuvio*0.2,
     vesuvio,
     vesuvio*2
   ), all=TRUE
 )
``` 

### Gamma correction

##### $y=x^{\gamma}$

```{r}
display(
   combine(
     vesuvio^0.7,
     vesuvio,
     vesuvio^2
   ), all=TRUE
 )
``` 

### Thresholding

Thresholding is a segmentation method that converts a gray-scale image into a binary image. During this operation all pixels whose intesity values are above a threshold are set to a foreground value, while all the remaining pixels are set to a background value.

#####  $y = x \le t$   or   $y = x \ge t$

```{r}
otsuGreyVesuvio = greyVesuvio < otsu(greyVesuvio)
print(otsuGreyVesuvio)
```

```{r}
display(
  combine(
    greyVesuvio,
    otsuGreyVesuvio
  ),all=TRUE
)
```

### Morphological operations

Morphological operations are usually applied on binary images.

**Erosion**: the value of the output pixel is the minimum value of all the pixels in the input pixel's neighborhood.<br>
**Dilation**: the value of the output pixel is the maximum value of all pixels in the input pixel's neighborhood.<br> 
**Opening**: is an erosion followed by a dilation.<br>
**Closing**: is a dilation followed by an erosion.

```{r}
kern=makeBrush(5,'diamond')
display(
   combine(
     otsuGreyVesuvio,
     erode(otsuGreyVesuvio,kern),
     dilate(otsuGreyVesuvio,kern),
     opening(otsuGreyVesuvio,kern),
     closing(otsuGreyVesuvio,kern)
   ), all=TRUE
)
```

`fillHull()` allows to fill holes in objects.

```{r}
display(
   combine(
     closing(otsuGreyVesuvio),
     fillHull(closing(otsuGreyVesuvio))
   ), all=TRUE
)
```

### Convolution

Convolution is a mathematical operation on two functions to produce a third function. Convolution can be written in the symbolic form as follows:

<center> 
$g(x,y)=(w*f)(x,y)=\sum_{i=-\infty}^{+\infty}\sum_{j=-\infty}^{+\infty}w(i,j)f(x+i,y+j)$
</center><br>

$g(x,y)$ is the filtered image, $f(x,y)$ is the original image, $w$ is the kernel.

```{r include = FALSE}
iPxs<-c();
oPxs<-c();
for (i in 1:81){
  iPxs<-c(iPxs,paste("$I_{",i,"}$",sep=""));  
  oPxs<-c(oPxs,paste("$O_{",i,"}$",sep=""));  
}
kPxs<-c();
for(i in 1:9){
  kPxs<-c(kPxs,paste("$K_{",i,"}$",sep=""));  
}
input=matrix(iPxs,nrow = 9,ncol = 9, byrow = TRUE);
kernel=matrix(kPxs,nrow = 3, ncol = 3, byrow = TRUE);
```

<center><div>
<div style="display:inline-block; padding:30px">
```{r echo=FALSE}
kable(input);
```
</div>

<div style="display:inline-block;  padding:30px; vertical-align:105px">
```{r echo=FALSE}
kable(kernel);
```
</div>
</div></center>

$O_{12}=I_{2}K_{1}+I_{3}K_{2}+I_{4}K_{3}+I_{11}K_{4}+I_{12}K_{5}+I_{13}K_{6}+I_{20}K_{7}+I_{21}K_{8}+I_{22}K_{9}$

#### Build kernel

##### Low-pass filter

A low-pass filter is used to blur images, remove artefacts. An example of low-pass filter is the gaussian filter. 
A filter can be generated using `makeBrush()`.

```{r}
gaussianKern=makeBrush(size = 5, shape = 'gaussian', sigma = 3)
```


```{r echo=FALSE}
kable(gaussianKern)
```

##### High-pass filter

A high-pass filter is used to detect edges, sharpen images. An example of high-pass filter is the laplacian filter.

```{r}
values=c(-1,0,-1,0,4,0,-1,0,-1)
laplacianKern=matrix(values,nrow = 3,ncol = 3, byrow = TRUE)
```

```{r echo=FALSE}
kable(laplacianKern)
```

##### Laplacian of gaussian filter

```{r}
values=c(0,0,-1,0,0,0,-1,-2,-1,0,-1,-2,16,-2,-1,0,-1,-2,-1,0,0,0,-1,0,0)
lapgauss5kern=matrix(values,nrow = 5,ncol = 5, byrow = TRUE)
```

```{r echo=FALSE}
kable(lapgauss5kern)
```

#### Filtering

`filter2()` allows to filter an image using the 2D FFT (2-Dimensional Fast Fourier Transform) convolution product.

```{r}
GaussImage=filter2(vesuvio,gaussianKern)
LaplacianImage=filter2(vesuvio,laplacianKern)
LoGImage=filter2(vesuvio,lapgauss5kern)
```

```{r}
display (
  combine(
    vesuvio,
    GaussImage,
    LaplacianImage,
    LoGImage
  ), all=TRUE    
)
```

```{r}
display (
  combine(
    channel(LaplacianImage,'grey'),
    channel(LoGImage,'grey')
  ), all=TRUE    
)
```

## Example: application in cell biology

"cells.tif" and "nuclei.tif" are fluorescent microscopy images from two channels of HeLa cells, distributed with EBImage.

```{r}
cells=readImage(system.file("images","cells.tif",package="EBImage"))
nuclei=readImage(system.file("images","nuclei.tif",package="EBImage"))
display(cells,all=TRUE)
display(nuclei,all=TRUE)

rgb = rgbImage(green=1.5*cells,blue=nuclei)
display(rgb,all=TRUE)
```

### Nuclei segmentation

```{r}
nmaskt = thresh(nuclei, w = 15, h = 15, offset = 0.05)
display(nmaskt,all=TRUE)
nmaskf = fillHull(opening(nmaskt, makeBrush(5, shape='disc')) )
display(nmaskf,all=TRUE)
```

```{r}
display( combine(getFrame(nmaskt, 3), getFrame(nmaskf, 3)), all=TRUE )
dmap = distmap(nmaskf)
display(dmap,all=TRUE)
range(dmap)
display(normalize(dmap), frame = 3)
nmask = watershed(dmap, tolerance = 2)
display(nmask,all=TRUE)
```

```{r}
display( combine(
  toRGB( getFrame(nuclei, 3) ), 
  colorLabels( getFrame(nmask, 3) )
), all=TRUE )
```

### Cytoplasm segmentation

```{r}
cmaskt = closing( gblur(cells, 1) > 0.105, makeBrush(5, shape='disc') )
```

```{r}
cmask  = propagate(cells, seeds=nmask, mask=cmaskt, lambda = 0.001)
```

```{r}
display( combine(
  toRGB( getFrame(cmaskt, 3) ), 
  colorLabels( getFrame(cmask, 3) )
), all=TRUE )
```

```{r}
display( paintObjects(nmask,
            paintObjects(cmask, rgb, col = "magenta", thick = TRUE),
         col = "yellow", thick = TRUE), all = TRUE)
```

```{r}
st = stackObjects(cmask, rgb)
display(st, all = TRUE)
```

```{r}
features=computeFeatures.shape(cmask[,,1], cells[,,1])
```

```{r}
head(features)
```

```{r echo=FALSE}
pch=21
color='black'
bg='#A2C353'
par(mfrow=c(2,3)) 
plot(features[,"s.area"],main='s.area',pch=pch,col=color,bg=bg)
plot(features[,"s.perimeter"],main='s.perimeter',pch=pch,col=color,bg=bg)
plot(features[,"s.radius.mean"],main='s.radius.mean',pch=pch,col=color,bg=bg)
plot(features[,"s.radius.sd"],main='s.radius.sd',pch=pch,col=color,bg=bg)
plot(features[,"s.radius.min"],main='s.radius.min',pch=pch,col=color,bg=bg)
plot(features[,"s.radius.max"],main='s.radius.max',pch=pch,col=color,bg=bg)
```

```{r echo=FALSE}
color='#448FA7'
par(mfrow=c(2,3)) 
hist(features[,'s.area'],main='Area',col=color)
hist(features[,"s.perimeter"],main='Perimeter',col=color)
hist(features[,"s.radius.mean"],main='Mean radius',col=color)
hist(features[,"s.radius.sd"],main='SD',col=color)
hist(features[,"s.radius.min"],main='Min radius',col=color)
hist(features[,"s.radius.max"],main='Max radius',col=color)
```

```{r echo=FALSE}
color='#942192'
lwd=3
par(mfrow=c(2,3)) 
plot(density(features[,"s.area"]),main='s.area',col=color,lwd=lwd)
plot(density(features[,"s.perimeter"]),main='s.perimeter',col=color,lwd=lwd)
plot(density(features[,"s.radius.mean"]),main='s.radius.mean',col=color,lwd=lwd)
plot(density(features[,"s.radius.sd"]),main='s.radius.sd',col=color,lwd=lwd)
plot(density(features[,"s.radius.min"]),main='s.radius.min',col=color,lwd=lwd)
plot(density(features[,"s.radius.max"]),main='s.radius.max',col=color,lwd=lwd)
```

## References

[Package 'EBImage'](https://www.bioconductor.org/packages/release/bioc/manuals/EBImage/man/EBImage.pdf)

[Introduction to EBImage](https://www.bioconductor.org/packages/devel/bioc/vignettes/EBImage/inst/doc/EBImage-introduction.html)

[Basics of image data and spatial patterns analysis in R](https://master.bioconductor.org/help/course-materials/2015/BioC2015/BioC2015Oles.html#26)